---
title: "SSA R Coding Workshop 2"
subtitle: "Data Cleaning & Manipulation With Dplyr"
author: "Andrew Bates"
output: ioslides_presentation
---

## This Workshop

> - `dplyr` is part of the [tidyverse](https://www.tidyverse.org):
>     - "The tidyverse is an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures."
> - "[dplyr](https://dplyr.tidyverse.org) is a grammar of data manipulation, providing a consistent set of verbs that help you solve the most common data manipulation challenges"
> - You **will** need to do data manipulation and cleaning
> - `dplyr` makes this easy and fun


## The Data 

> - Reviews of burritos in San Diego
> - Project of [Scott Cole](https://srcole.github.io/100burritos/), a neuroscience student at UCSD
> - Find it
>     - on his [website](https://github.com/asbates/ssa_r_workshops/tree/master/fall_18/data)
>     - or [Kaggle](https://www.kaggle.com/srcole/burritos-in-san-diego/version/2#burritos_01022018.csv)


<div class="notes">
- How to follow along?
    1. Type the code shown on the slides
    2. Click on the green arrow in the code chunk of the source file
    3. Copy the code in the source file into the console
- Start by typing out the code
- If that starts taking too long, then run the code chunk
- The purpose is to get a feel for the package
- We don't want to spend too much time trying to fix bugs caused by typing something incorrectly
</div>


## Setup

1. Go to https://github.com/asbates/ssa_r_workshops
2. Click the green "Clone or download" button
3. Select "Download ZIP" (unless you're familiar with Git)
4. Unzip and click on `ssa_r_workshops.Rproj` 
5. Install (if needed) and load the packages we will use
6. Read in the data

```{r setup_shown, eval = FALSE}
install.packages("dplyr", "readr", "here") 
library(dplyr)
library(readr)
library(here)
dirty_burrito <- read_csv(here("fall_18", "data", "burritos_01022018.csv"))
# or, if you don't want to use the here package
#dirty_burrito <- read_csv("path/to/file/burritos_01022018.csv")
```
     
```{r setup_hidden, echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(readr)
library(here)
dirty_burrito <- read_csv(here("fall_18", "data", "burritos_01022018.csv"))

```



## First Look

```{r first_look_print}
dirty_burrito
```

## First Look
```{r first_look_names}
names(dirty_burrito)
```


## Some Issues

> - There are problems with some names:
>     - `readr` already took care of one issue (2 `Salsa` columns)
>     - Spaces in names are not very R friendly (`Mass (g)`)
>     - `Bell peper` is spelled wrong
>     - Inconsistent naming scheme (capitalization)
> - There are some `NA` values (not necessarily a problem)
> - `Chips` encoding is not so obvious (does `x` mean yes?)
> - What's the difference between `Tortilla` and `Wrap`?
> - The last two can probably be figured out with a data dictionary



## Fixing The Issues

> - To simplify things, let's start by:
>     - `select`ing just a few variables to work with
>     - and `rename`ing them

## Selecting Variables of Interest

```{r select_vars}
clean_burrito <- dirty_burrito %>%
  select(Location,
         Burrito,
         Yelp,
         Google,
         Cost,
         `Mass (g)`,
         Tortilla,
         Meat,
         Synergy,
         overall,
         Beans)
```

<div class="notes">
Note the use of backticks ` `` ` to handle the space in Mass
</div>

## Selecting Variables of Interest
```{r clean_first_pass}
clean_burrito
```


## Rename Variables

```{r rename}
clean_burrito <- clean_burrito %>%
  rename(
    location = Location,
    burrito = Burrito,
    yelp_rating = Yelp,
    google_rating = Google,
    cost = Cost,
    mass = `Mass (g)`,
    tortilla = Tortilla,
    meat = Meat,
    synergy = Synergy,
    overall_rating = overall,
    beans = Beans
  )
```

<div class="notes">
We rename using `new_name = old_name`
</div>

## Rename Variables

```{r}
clean_burrito
```

## Quiz 1

> - What does `select` do?
>     - `select`s variables (columns)
> - What does `rename` do?
>     - This should be obvious
> - What's this `%>%` thing?
>     - The 'pipe' operator
>     - From the package `magrittr`. Comes with `dplyr`
>     - Improves readability of code
>         - `x %>% h %>% g %>% f` instead of `f(g(h(x)))`
>         - "Take x. Apply the function h. Then g. Then f."
>     - First argument is always the data
>     - Allows for easy chaining of operations on data frame


<div class="notes">
- It might seem a bit weird at first but you get used to it in time
- A lot of things are easier to read & write than in base R
</div>

## Recode Beans

 - Let's recode `beans`
 - We'll assume that an `x` means the burrito has beans
 - And `NA` means there are no beans
 - But first, let's see if `beans` takes any other values

```{r beans-levels}
clean_burrito %>%
  pull(beans) %>%
  as.factor() %>%
  levels
```

- OK. So `beans` can be either big or little x.



## Recode Beans

- Let's set `beans` to "yes" if it's either "x" or "X" and "no" otherwise
- We do this by `mutate`ing the original variable

```{r beans-yesno}
clean_burrito <- clean_burrito %>%
  mutate(
    beans = if_else(beans == "x" | beans == "X", "yes", "no")
  )
```




## Recode Beans

- Let's check to make sure we got what we wanted

```{r beans-yesno-check}
clean_burrito %>%
  select(beans)
```


## Recode Beans

- Whoops! It looks like we didn't replace the `NA` values.
- Let's try again

```{r beans-yesno-take2}
clean_burrito <- clean_burrito %>%
  mutate(
    beans = if_else(is.na(beans), "no", beans)
  )
```


<div class="notes">
- We could also do `clean_burrito <- clean_burrito %>% mutate(beans = ifelse(is.na(beans), "no", "yes"))` in one step
- In practice, you might do this to clean it up
</div>

## Recode Beans

- And check our work

```{r beans-yesno-check2}
clean_burrito %>%
  select(beans)
```


## Quiz 2


> - What's the difference between `select` and `pull`?
>     - `pull` gives us the *vector* `beans`
>     - `select` gives us a *data frame* with the column `beans`
> - What does `if_else()` do?
>     - If the first argument is true (`beans` is "x" or `beans` is "X)
>     - Do the second argument (set to "yes")
>     - Otherwise, do the third argument (set to "no")


<div class="notes">

- Imagine we have a bag of candy
- If we `select` Skittles, we get back a bag with only Skittles
- If we `pull` Skittles, we get back Skittles without the bag

</div>



## Summarize Overall Rating

- Great success!
- Let's take a closer look by `summarise`ing the data.
- We'll start with `overall_rating`

```{r overll-summ}
clean_burrito %>%
  summarise(
    min = min(overall_rating),
    mean = mean(overall_rating),
    median = median(overall_rating),
    max = max(overall_rating)
  )
```


## Summarize Overall Rating

- Uh oh! It looks like some burritos don't have ratings

```{r overall-summ2}
clean_burrito %>%
  summarise(
    min = min(overall_rating, na.rm = TRUE),
    mean = mean(overall_rating, na.rm = TRUE),
    median = median(overall_rating, na.rm = TRUE),
    max = max(overall_rating, na.rm = TRUE)
  )
```

## Summarize Rating By Burrito Type


```{r overall-group-by-typ}
clean_burrito %>%
  group_by(burrito) %>%
  summarise(
    mean_rating = mean(overall_rating, na.rm = TRUE)
  ) %>% 
  top_n(5)
```



## Summarize Rating By Location


```{r overall-group-by-loc}
clean_burrito %>%
  group_by(location) %>%
  summarise(
    mean_rating = mean(overall_rating, na.rm = TRUE)
  ) %>% 
  top_n(5)
```


## Average Rating for Burritos with Beans

```{r overall-beans}
clean_burrito %>%
  filter(beans == "yes") %>%
  group_by(location) %>%
  summarise(
    mean_rating = mean(overall_rating, na.rm = TRUE)
  ) %>% 
  top_n(-3)
```


## Quiz 3

> - What did we learn?
>     - California-like burritos are the way to go
>     - Get one at Chili Peppers
>     - Don't go to Goody's if you like beans


<div class="notes">

- A couple of times, we ran some code, didn't like the output, and ran some more
- Why?
- I wanted to emulate what your workflow would be like as much as possible
- Not everything will work out the way you wanted the first time
- You have to look at the result and change things as necessary
- This is just how writing code works
- As we went along, we add more `%>%`s to each thing we did
- Until you get more familiar with `%>%`, you should just do one step at a time, check the output, then add another step
</div>


## Next Steps

- Practice, practice, practice!
- Do something similar with a different/bigger subset or even the full data set
- Use R Markdown to write about what you do
- Look into other functions in `dplyr`
- Teach your friends



